---
title: "Prepare Data"
output: html_document
date: '2023-07-28'
---

```{r setup, include=FALSE}
source("../in_progress/findDomain.R")
source("../in_progress/prepareSFE.R")
source("../in_progress/wrap_unwrap_SFE_raster.R")

library(Voyager)
library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(SingleCellExperiment)
library(ggplot2)
library(scater)
#> Loading required package: scuttle
library(scuttle)
library(scran)
library(stringr)
library(patchwork)
library(bluster)
library(rjson)
library(terra)
theme_set(theme_bw())

```


## Sample Metadata

The data being used in this workflow is taken from "Spatial Gene Expression 
Changes in the Mouse Heart After Base-Targeted Irradiation"

Radiation cardiotoxicity (RC) is a clinically significant 
adverse effect of treatment for patients with thoracic malignancies. 
Clinical studies in lung cancer have indicated that heart substructures 
are not uniformly radiosensitive, and that dose to the heart base drives RC.

This was a pilot experiment to determine whether spatial transcriptomics 
could provide useful insights in this context. 


### Experimental Design:
2 samples
2 conditions - irradiated heart and control
4 adjacent sections taken from each tissue block (sample)
8 sections in total


The sample metadata is summarised below:


```{r}
dataDir <- "../testData/results"

samples <- list.files( dataDir)


samples_short <- samples %>% 
  gsub("results_", "", .)  %>% 
  gsub("_Results", "", .)

sample_id <- samples_short %>% 
  gsub("_.*", "", .)  



condition <- c(rep("Control", 4), rep("irradiated", 4))
tissue <- c(c(rep("tissue1", 4), rep("tissue2", 4)))

sampleInfo <- data.frame(Fname = samples, 
          id = samples_short, 
          condition, 
          tissue, 
          sample_id
          )

sampleInfo


pathToTenXOuts <- file.path(dataDir, sampleInfo$Fname, "outs")
```

## Import the expression data

The Voyager library Load in data and add in QC metrics

More info at voyager


```{r}


sfe <- SpatialFeatureExperiment::read10xVisiumSFE(
                                      dirs = pathToTenXOuts, 
                                      samples = sampleInfo$Fname, 
                                      sample_id = sampleInfo$sample_id, 
                                      type = "sparse", 
                                      data = "filtered", 
                                      images = "hires",
                                      zero.policy=TRUE)



sfe_single <- SpatialFeatureExperiment::read10xVisiumSFE(
                                      dirs = pathToTenXOuts, 
                                      samples = sampleInfo$Fname[1], 
                                      sample_id = sampleInfo$sample_id[1], 
                                      type = "sparse", 
                                      data = "filtered", 
                                      images = "hires",
                                      zero.policy=TRUE)


```





## Add metadata and spot positions to data

```{r}
sfe <- prepareSFE(sfe, sampleInfo)
```


## Add Image annotations

```{r}
annotationDir <- "C:/Users/nrq2/OneDrive - Newcastle University/SpanielTest/testData/annotation_images"
annotations <- list.files(annotationDir) 
annoInfo <- data.frame(sample_id = gsub("_.*", "", annotations) ,
  domain = gsub(".*_", "", annotations) %>% gsub("\\.jpg", "", .),jpg = annotations) %>% dplyr::filter(domain != "great-vessels")




annoInfo
```


```{r}
## to fix
#sfe <- addAnnotations(sfe, annoInfo, annotationDir)
##to fix! - doesn't work for great-vessels
## factor the domain column so that domain is plotted on top



sfe <- findAllDomains(annoInfo, sfe)

domains <- annoInfo$domain[annoInfo$sample_id == "C01"]
```

## replace with spaniel plot images
```{r}
p1 <- plotSpatialFeature(sfe, domains[1:4],
                   image_id = "hires", 
                   maxcell = 5e4, 
                   ncol = 2, 
                   sample_id = "C01", linetype = 0)
p2 <- plotSpatialFeature(sfe, domains[5:8],
                   image_id = "hires", 
                   maxcell = 5e4, 
                   ncol = 2, 
                   sample_id = "C01", linetype = 0)

p1
p2

```

```{r}

```




```{r}
is_mt <- str_detect(rowData(sfe)$symbol, "^mt-")
sfe <- addPerCellQCMetrics(sfe, subsets = list(mito = is_mt))
```


Spots where no genes are detected can be removed from the remainder of the 
analysis.

```{r}
filter <- sfe$detected > 0
sfe <- sfe[, filter]
```


The filtered data can then be normalised using the the "logNormCounts" 
function from scuttle  
and the expression of selected genes can be viewed on the histological image.




```{r}
sfe <- scuttle::logNormCounts(sfe)

dec <- scran::modelGeneVar(sfe)
hvgs <- getTopHVGs(dec, prop=0.1)


sfe <- scater::runPCA(sfe,
                      ncomponents = 30,
                      scale = TRUE, 
                      subset_row = hvgs)

sfe <- scater::runUMAP(sfe, dimred="PCA")


clust <- scran::clusterCells(sfe, BLUSPARAM=NNGraphParam(cluster.fun="louvain"), 
                         use.dimred="PCA", full=TRUE)

sfe$clust <- clust$clusters



```


```{r}
scater::plotUMAP(sfe, colour_by = "left-ventricle")
scater::plotUMAP(sfe, colour_by = "clust")
```




### TO Do! function to make

compare clusters with histological region, number of spots, consistency etc

## Wrap and Save Data

```{r}
sfe_wrapped <- wrapSFE(sfe)
saveRDS(sfe_wrapped, "../testData/rObjects/wrapped_sfe.rds")
```


## Unwrap Data

```{r}
sfe_wrapped <- readRDS("../testData/rObjects/wrapped_sfe.rds")
sfe <- unwrapSFE(sfe_wrapped)

## check it loads
p1 <- plotSpatialFeature(sfe, features = "domain",
                   image_id = "hires", 
                   maxcell = 5e4, 
                   ncol = 2, 
                   sample_id = "C01",
                   color = "black", linetype = 0,
                   
                   ) + theme_classic()
p1


```

